---
title: "POLS0066 - Data Analysis - Ryan"
author: "GNWL7"
date: "2024-01-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Affective Polarisation in the French multiparty system\`

## Setup

The data comes from Sciences Po's [Dataverse](https://data.sciencespo.fr/dataverse/donneesrecherche). Data from Waves 1-4 has been imported as csv's, and stored in the /data folder. It has been aggregated and cleaned in `data_cleaning.R`.

```{r include=FALSE}
library(tidyverse)
library(viridis)
library(lmtest)
library(multiwayvcov)
library(lme4)
library(sjPlot)
load("data/recoded_data.RData")
```

### Adding Party Vote Share

In- and out-group evaluations will be weighted using vote shares so that larger parties count more. We use data for the first round of the 2022 Presidential election, from the French Ministry of the Interior (Ministère de l'Intérieur). Although this proxy has limitations, it has been used in previous studies and constitutes an improvement from unweighted evaluations.

In the calculations of in- and out-bloc sympathy, each party's evaluation will be weighted by the ratio of votes gained by that party as a part of the bloc's overall votes. See Reiljan & Ryan (2021, Appendix) for more details.

```{r}
party_vote_shares <- tribble(
  ~party, ~vote_share, ~bloc, ~relative_vote_share,
  "LFI", 22.0, "left_greens", round(22.0 / (22.0 + 4.6 + 1.8),2),
  "EELV", 4.6, "left_greens", round(4.6 / (22.0 + 4.6 + 1.8),2),
  "PS", 1.8, "left_greens", round(1.8 / (22.0 + 4.6 + 1.8), 2),
  "LREM", 27.9, "centre_right", round(27.9 / (27.9 + 4.8),2),
  "LR", 4.8, "centre_right", round(4.8 / (27.9 + 4.8),2),
  "Reconquete", 7.1, "far_right", round(7.1 / (7.1 + 23.2),2),
  "RN", 23.2, "far_right", round(23.2 / (7.1 + 23.2), 2)
)
```


## Waves 2, 3 & 4 Analysis - The Ideological Drivers of Affective polarisation Within the Three Blocs

First supporters within the Left-plus-Greens bloc:

```{r}
left_greens <- c("LFI", "EELV", "PS")
centre_right <- c("LREM", "RN")
far_right <- c("RN", "Reconquete")

# Subset respondents of waves 1, 3 & 4
df_W134 <- aggreg_waves_recoded |> 
  filter(W1 & W3 & W4 & in_party_sympathy & left_right_self)


# ---- Left-plus-Greens bloc ---- 
left_bloc_df <- aggreg_waves_recoded |> 
  filter(!is.na(in_party_sympathy) 
         & PID %in% left_greens) |> 
  
  # out-group and in-group sympathy, weighted by vote share
  mutate(
    out_group_sympathy_centre_right = 
      1/2 * group_sentiment_LREM * party_vote_shares$relative_vote_share[party_vote_shares$party == "LREM" ] +
      1/2 * group_sentiment_LR * party_vote_shares$relative_vote_share[party_vote_shares$party == "LR" ] ,
    out_group_sympathy_far_right = 
      1/2 * group_sentiment_Reconquete * party_vote_shares$relative_vote_share[party_vote_shares$party == "Reconquete" ] +
      1/2 * group_sentiment_RN * party_vote_shares$relative_vote_share[party_vote_shares$party == "RN" ] ,
    in_group_sympathy = 
      1/3 * group_sentiment_LFI * party_vote_shares$relative_vote_share[party_vote_shares$party == "LFI" ] +
      1/3 * group_sentiment_EELV * party_vote_shares$relative_vote_share[party_vote_shares$party == "EELV" ] +
      1/3 * group_sentiment_PS * party_vote_shares$relative_vote_share[party_vote_shares$party == "PS" ]
    ) |> 
  
  # affective polarisation
  mutate(affect_polarisation_centre_right = in_party_sympathy - out_group_sympathy_centre_right,
         affect_polarisation_far_right = in_party_sympathy - out_group_sympathy_far_right)

# ---- Models ----
# Polarisation with centre-right
model_left_and_centre_right <- lm(
  data = left_bloc_df, affect_polarisation_centre_right ~ 
    PID_strength + socioeco + in_group_sympathy + environment +
    + immigration + interest_politics
  )
summary(model_left_and_centre_right)

# Polarisation with far-right
model_left_and_far_right <- lm(
  data = left_bloc_df, affect_polarisation_far_right ~ 
    PID_strength + socioeco + in_group_sympathy + environment +
    + immigration + interest_politics
  )
summary(model_left_and_far_right)

```

Next, respondents within the centre-right group. Importantly, there are less respondents within this bloc: N = 149.

```{r}
# ---- Centre-right bloc ---- 
far_right_bloc_df <- aggreg_waves_recoded |> 
  filter(!is.na(in_party_sympathy) 
         & PID %in% far_right) |> 
  
  # out-group and in-group sympathy, weighted by vote share
  mutate(
    out_group_sympathy_left = 
      1/3 * group_sentiment_LFI * party_vote_shares$relative_vote_share[party_vote_shares$party == "LFI" ] +
      1/3 * group_sentiment_EELV * party_vote_shares$relative_vote_share[party_vote_shares$party == "EELV" ] +
      1/3 * group_sentiment_PS * party_vote_shares$relative_vote_share[party_vote_shares$party == "PS" ],
    out_group_sympathy_center_right = 
      1/2 * group_sentiment_LREM * party_vote_shares$relative_vote_share[party_vote_shares$party == "LREM" ] +
      1/2 * group_sentiment_LR * party_vote_shares$relative_vote_share[party_vote_shares$party == "LR" ] ,
    in_group_sympathy = 
      1/2 * group_sentiment_Reconquete * party_vote_shares$relative_vote_share[party_vote_shares$party == "Reconquete" ] +
      1/2 * group_sentiment_RN * party_vote_shares$relative_vote_share[party_vote_shares$party == "RN" ]
    ) |> 
  
  # affective polarisation
  mutate(affect_polarisation_left = in_party_sympathy - out_group_sympathy_left,
         affect_polarisation_center_right = in_party_sympathy - out_group_sympathy_center_right)

# ---- Models ----
# Polarisation with left-plus-greens
model_left_and_far_right <- lm(
  data = far_right_bloc_df, affect_polarisation_left ~ 
    PID_strength + socioeco + in_group_sympathy + environment +
    + immigration + interest_politics
  )
summary(model_left_and_far_right)

# Polarisation with far-right
model_far_and_center_right <- lm(
  data = far_right_bloc_df, affect_polarisation_center_right ~ 
    PID_strength + socioeco + in_group_sympathy + environment +
    + immigration + interest_politics
  )
summary(model_far_and_center_right)
```

Finally, let's observe the determinents of affective polarisation for supporters of parties within the far-right bloc:

```{r}
# ---- Far-Right group ---- 
centre_right_bloc_df <- aggreg_waves_recoded |> 
  filter(!is.na(in_party_sympathy) 
         & PID %in% centre_right) |> 
  
  # out-group and in-group sympathy, weighted by vote share
  mutate(
    out_group_sympathy_left = 
      1/3 * group_sentiment_LFI * party_vote_shares$relative_vote_share[party_vote_shares$party == "LFI" ] +
      1/3 * group_sentiment_EELV * party_vote_shares$relative_vote_share[party_vote_shares$party == "EELV" ] +
      1/3 * group_sentiment_PS * party_vote_shares$relative_vote_share[party_vote_shares$party == "PS" ],
    out_group_sympathy_far_right = 
      1/2 * group_sentiment_Reconquete * party_vote_shares$relative_vote_share[party_vote_shares$party == "Reconquete" ] +
      1/2 * group_sentiment_RN * party_vote_shares$relative_vote_share[party_vote_shares$party == "RN" ] ,
    in_group_sympathy = 
      1/2 * group_sentiment_LREM * party_vote_shares$relative_vote_share[party_vote_shares$party == "LREM" ] +
      1/2 * group_sentiment_LR * party_vote_shares$relative_vote_share[party_vote_shares$party == "LR" ]
    ) |> 
  
  # affective polarisation
  mutate(affect_polarisation_left = in_party_sympathy - out_group_sympathy_left,
         affect_polarisation_far_right = in_party_sympathy - out_group_sympathy_far_right)

# ---- Models ----
# Polarisation with left-plus-greens
model_centre_right_and_left <- lm(
  data = centre_right_bloc_df, affect_polarisation_left ~ 
    PID_strength + socioeco + in_group_sympathy + environment +
    + immigration + interest_politics
  )
summary(model_centre_right_and_left)

# Polarisation with centre-right
model_centre_and_far_right <- lm(
  data = centre_right_bloc_df, affect_polarisation_far_right ~ 
    PID_strength + socioeco + in_group_sympathy + environment +
    + immigration + interest_politics
  )
summary(model_centre_and_far_right)
```
