---
title: "POLS0066 - Data Analysis - Ryan"
author: "GNWL7"
date: "2024-01-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Affective Polarisation in the French multiparty system\`

## Setup

The data comes from Sciences Po's [Dataverse](https://data.sciencespo.fr/dataverse/donneesrecherche). Data from Waves 1-4 has been imported as csv's, and stored in the /data folder. It has been aggregated and cleaned in `data_cleaning.R`.

```{r include=FALSE}
library(tidyverse)
library(viridis)
library(lmtest)
library(multiwayvcov)
library(lme4)
library(sjPlot)
load("data/recoded_data.RData")
```

## Wave 4 Analysis - The Impact of Ideological Difference and Partisan Identity on Affective Polarisation

We can now move on to the testing of our hypotheses.

The fourth wave of the 2022 French Election Study, carried out after the Presidential election from April 28 to June 2, 2022, takes up the entirety of module 6 (2021-2026) of the CSES (Comparative study of electoral systems) international comparison project.

```{r}
# Subset participants who:
#   participated in wave 4
#   have expressed an in-party sympathy
#   have revealed their position on the left-right scale
# N = 1034
df_W4 <- aggreg_waves_recoded |> 
  filter(W4 & in_party_sympathy & left_right_self)
```

### Adding Ideological distance

To include ideological distance in models, both the objective and subjective ideological positions of parties can be computed. Objective ideological position is the respondents' average rating of the party's left-right position. Subjective ideological position is each respondent's rating of the party's left-right position.

```{r}
df_ideological <- df_W4 |>
  mutate(
    subj_ideo_dist_LFI = abs(left_right_LFI - left_right_self),
    subj_ideo_dist_EELV = abs(left_right_EELV - left_right_self),
    subj_ideo_dist_PS = abs(left_right_PS - left_right_self),
    subj_ideo_dist_MoDem = abs(left_right_MoDem - left_right_self),
    subj_ideo_dist_LREM = abs(left_right_LREM - left_right_self),
    subj_ideo_dist_LR = abs(left_right_LR - left_right_self),
    subj_ideo_dist_RN = abs(left_right_RN - left_right_self),
    subj_ideo_dist_Reconquete = abs(left_right_Reconquete - left_right_self)
  )|> 
  mutate(
    obj_ideo_dist_LFI = abs(mean(left_right_LFI, na.rm = TRUE) - left_right_self),
    obj_ideo_dist_EELV = abs(mean(left_right_EELV, na.rm = TRUE) - left_right_self),
    obj_ideo_dist_PS = abs(mean(left_right_PS, na.rm = TRUE) - left_right_self),
    obj_ideo_dist_MoDem = abs(mean(left_right_MoDem, na.rm = TRUE) - left_right_self),
    obj_ideo_dist_LREM = abs(mean(left_right_LREM, na.rm = TRUE) - left_right_self),
    obj_ideo_dist_LR = abs(mean(left_right_LR, na.rm = TRUE) - left_right_self),
    obj_ideo_dist_RN = abs(mean(left_right_RN, na.rm = TRUE) - left_right_self),
    obj_ideo_dist_Reconquete = abs(mean(left_right_Reconquete, na.rm = TRUE) - left_right_self)
  ) |> 
  mutate(
    ideo_extreme = abs(left_right_self - 5)
  ) |> 
  select(-left_right_LFI, -left_right_EELV, -left_right_PS, -left_right_LREM,
         -left_right_LR, -left_right_RN, -left_right_Reconquete, -left_right_MoDem)

```

Starting with subjective ideological distance:

```{r}
# Long dataset with, for each respondent, their subjective ideological distance
# and party symapthy for all parties
df_party_dislike_subj_long <- 
  df_ideological |> 
  select(
    id, age_decade, diploma, sex,
    party_sympathy_LFI, party_sympathy_EELV, party_sympathy_PS,
    party_sympathy_LREM, party_sympathy_LR, party_sympathy_RN,
    party_sympathy_Reconquete, party_sympathy_MoDem,
    interest_politics, PID, PID_strength, ideo_extreme, 
    subj_ideo_dist_LFI, subj_ideo_dist_EELV, subj_ideo_dist_PS,
    subj_ideo_dist_MoDem, subj_ideo_dist_LREM, subj_ideo_dist_LR,
    subj_ideo_dist_RN, subj_ideo_dist_Reconquete) |> 
  pivot_longer(
    cols = starts_with("party_sympathy_"),
    names_to = "party",
    values_to = "party_sympathy"
  ) |> 
  pivot_longer(
    cols = starts_with("subj_ideo_dist_"),
    names_to = "subj_ideo_dist",
    values_to = "subj_ideo_distance"
  ) |> 
  mutate(party = gsub("party_sympathy_", "", party)) |> 
  filter(party == gsub("subj_ideo_dist_", "", subj_ideo_dist)) |> 
  filter(PID != party)
```

Party sympathy for out-parties can be regressed on subjective ideological distance. We use multilevel linear regression models because the data structure is nested within the respondent, with 7 dyads between the respondent and the out-party.

```{r}
# Model 1: Subjective Ideological Distance, Political Interest, and Demographic controls
# Multilevel linear regression model with random intercepts
model1_party_dislike_subj <- lmer(
  data = df_party_dislike_subj_long,
  party_sympathy ~ 
    subj_ideo_distance + ideo_extreme + interest_politics + age_decade + diploma + sex + (1 | id)
)

tab_model(model1_party_dislike_subj)


# Model 2: Adding partisan identity strength
model2_party_dislike_subj <- lmer(
  data = df_party_dislike_subj_long,
  party_sympathy ~ 
    subj_ideo_distance + PID_strength + ideo_extreme + interest_politics + age_decade + diploma + sex + (1 | id)
)

tab_model(model2_party_dislike_subj)


# Model 3: Interaction between subjective ideological difference and political identity strength 
model3_party_dislike_subj <- lmer(
  data = df_party_dislike_subj_long,
  party_sympathy ~ 
    subj_ideo_distance*PID_strength + ideo_extreme + interest_politics + age_decade + diploma + sex + (1 | id)
) 

tab_model(model3_party_dislike_subj)

# Model 4: Interaction between subjective ideological difference and ideological extremity
model4_party_dislike_subj <- lmer(
  data = df_party_dislike_subj_long,
  party_sympathy ~ 
    subj_ideo_distance*ideo_extreme + PID_strength + interest_politics + age_decade + diploma + sex + (1 | id)
) 

tab_model(model4_party_dislike_subj)
```

Of course, using the subjective measure of ideological distance does not fully exclude the possibility that the relationship may partly work reversed as well, and that a more negative affect towards a group may also cause people to perceive these groups as more ideologically distant. For this reason, we also run the same model with the objective ideological distance measure instead" (Erkel & Turkenburg, 2023, p. 396)

```{r}
# Long dataset with, for each respondent, their objective ideological distance
# and party symapthy for all parties
df_party_dislike_obj_long <- 
  df_ideological |> 
  select(
    id, age_decade, diploma, sex,
    party_sympathy_LFI, party_sympathy_EELV, party_sympathy_PS,
    party_sympathy_LREM, party_sympathy_LR, party_sympathy_RN,
    party_sympathy_Reconquete, party_sympathy_MoDem,
    interest_politics, PID, PID_strength, ideo_extreme, 
    obj_ideo_dist_LFI, obj_ideo_dist_EELV, obj_ideo_dist_PS,
    obj_ideo_dist_MoDem, obj_ideo_dist_LREM, obj_ideo_dist_LR,
    obj_ideo_dist_RN, obj_ideo_dist_Reconquete) |> 
  pivot_longer(
    cols = starts_with("party_sympathy_"),
    names_to = "party",
    values_to = "party_sympathy"
  ) |> 
  pivot_longer(
    cols = starts_with("obj_ideo_dist_"),
    names_to = "obj_ideo_dist",
    values_to = "obj_ideo_distance"
  ) |> 
  mutate(party = gsub("party_sympathy_", "", party)) |> 
  filter(party == gsub("obj_ideo_dist_", "", obj_ideo_dist)) |> 
  filter(PID != party)
```

On with the multilevel linear regression models:

```{r}
# Model 1: Objective Ideological Distance, Political Interest, and Demographic controls
# Multilevel linear regression model with random intercepts
model1_party_dislike_obj <- lmer(
  data = df_party_dislike_obj_long,
  party_sympathy ~ 
    obj_ideo_distance + ideo_extreme + interest_politics + age_decade + diploma + sex + (1 | id)
)

tab_model(model1_party_dislike_obj)


# Model 2: Adding partisan identity strength
model2_party_dislike_obj <- lmer(
  data = df_party_dislike_obj_long,
  party_sympathy ~ 
    obj_ideo_distance + PID_strength + ideo_extreme + interest_politics + age_decade + diploma + sex + (1 | id)
)

tab_model(model2_party_dislike_obj)


# Model 3: Interaction between objective ideological difference and political identity strength 
model3_party_dislike_obj <- lmer(
  data = df_party_dislike_obj_long,
  party_sympathy ~ 
    obj_ideo_distance*PID_strength + ideo_extreme + interest_politics + age_decade + diploma + sex + (1 | id)
) 

tab_model(model3_party_dislike_obj)

# Model 4: Interaction between objective ideological difference and ideological extremity
model4_party_dislike_obj <- lmer(
  data = df_party_dislike_obj_long,
  party_sympathy ~ 
    obj_ideo_distance*ideo_extreme + PID_strength + interest_politics + age_decade + diploma + sex + (1 | id)
) 

tab_model(model4_party_dislike_obj)
```
